---
title: "ML with R - KNN Project"
author: "Suhas. P. K"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    default_style: "dark"
    downcute_theme: "chaos"
    toc: 6
---


```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE)
```

## K Nearest Neighbour Project

Since KNN is such a simple algorithm, we will just use this "Project" as a simple exercise to test your understanding of the implementation of KNN. 

## `Iris`  Dataset

We'll use the famous [iris](https://en.wikipedia.org/wiki/Iris_flower_data_set) data set for this project. It's a small data set with flower features that can be used to attempt to predict the species of an iris flower.

```{r libraries, message=FALSE,warning=FALSE}
if(!require(ggpubr)){
  install.packages("ggpubr")
  library(ggpubr)
}

if(!require(ISLR2)){
  install.packages("ISLR2")
  library(ISLR2)
}

if(!require(ggplot2)){
  install.packages("ggplot2")
  library(ggplot2)
}

if(!require(ggdark)){
  install.packages("ggdark")
  library(ggdark)
}

if(!require(caTools)){
  install.packages("caTools")
  library(caTools)
}

if(!require(class)){
  install.packages("class")
  library(class)
}

if(!require(plotly)){
  install.packages("plotly")
  library(plotly)
}




```

### Loading `iris` dataset
```{r loading dataset, warning=FALSE,message=FALSE}
df <- iris
head(df)
```

- Dataset structure

```{r dataset structure}
str(df)
```

- Dataset summary

```{r dataset summary, warning=FALSE,message=FALSE}
summary(df)
```

### Standardize data

- The iris data set has all its features in the same order of magnitude, but its good practice (especially with KNN) to standardize features in your data.

**Use `scale()` to standardize the feature columns of the iris dataset. Set this standardized version of the data as a new variable.**

```{r standardize data}
standard.feature <- scale( df[1:4])
```

- Checking that the scaling worked by checking the variance of one of the new columns.

```{r variance}
var(standard.feature[,1])
```

- Join the standardized data with the response/target/label column (the column with the species names.

```{r final dataset, warning=FALSE,message=FALSE}
final.data <- cbind(standard.feature,df[5])
head(final.data)
```

## Train-test split

```{r test-train split, message=FALSE, warning=FALSE}
set.seed(101)

sample <- sample.split(final.data$Species, SplitRatio = 0.70)
train <- subset(final.data, sample = TRUE)
test <- subset(final.data, sample = FALSE)
```


## Build a KNN model.

```{r knn plotting}
# Fit KNN model
k <- 10
knn_model <- knn(train[, 1:2], test[, 1:2], train[, 5], k = k)

# Combine predictions with test data
test_with_pred <- cbind(test, Predicted_Species = knn_model)

# Plotting using ggplot2 and ggpubr
ggplot(test_with_pred, aes(x = Sepal.Width, y = Sepal.Length, color = Predicted_Species)) +
  geom_point(size = 3, alpha = 0.9) +
  stat_ellipse(aes(fill = Predicted_Species, color = Predicted_Species), geom = "polygon", alpha = 0.4,color = "white") +
  labs(title = paste("KNN Cluster Probability Ellipses (k = ", k, ")", sep = ""),
       x = "Sepal.Length", y = "Sepal.Width", fill = "Predicted Species", color = "Predicted Species") +
  scale_color_manual(values = c("setosa" = "blue", "versicolor" = "green", "virginica" = "red")) +
  scale_fill_manual(values = c("setosa" = "lightblue", "versicolor" = "lightgreen", "virginica" = "lightcoral")) +
  dark_theme_light()


```


```{r}


# Fit KNN model
k <- 5
knn_model <- knn(train[, 1:2], test[, 1:2], train[, 5], k = k)

# Combine predictions with test data
test_with_pred <- cbind(test, Predicted_Species = knn_model)

# Plotting using ggplot2 and ggpubr
knn.plt <- ggplot() +
  geom_point(data = train, aes(x = Sepal.Width, y = Sepal.Length, color = "Train"), size = 3, alpha = 0.7) +
  geom_point(data = test_with_pred, aes(x = Sepal.Width, y = Sepal.Length, color = Predicted_Species), size = 3, alpha = 0.7) +
  stat_ellipse(data = train, aes(x = Sepal.Width, y = Sepal.Length, fill = "Train"), geom = "polygon", alpha = 0.2, color = "black") +
  stat_ellipse(data = test_with_pred, aes(x = Sepal.Width, y = Sepal.Length, fill = Predicted_Species), geom = "polygon", alpha = 0.2, color = "black") +
  labs(title = paste("KNN Cluster Probability Ellipses (k =", k, ")", sep = ""),
       x = "Sepal.Length", y = "Sepal.Width", fill = "Dataset") +
  scale_color_manual(values = c("Train" = "blue", "setosa" = "green", "versicolor" = "yellow", "virginica" = "red")) +
  scale_fill_manual(values = c("Train" = "lightblue", "setosa" = "lightgreen", "versicolor" = "lightcoral", "virginica" = "lightyellow")) +
  dark_theme_light()

ggplty <- ggplotly(knn.plt) 

ggplty

```



























































































































